Permutation Pattern matching in\\ $(213,231)$-avoiding permutations

### 2016-10-13 article.tex
revision

@@

For an occurrence $s$ of $\sigma$ in $\pi$,
we say that $\pi[j]$ correspond $\sigma[i]$ in the occurrence or
that $\sigma[i]$ is matched to $\pi[j]$  in the occurrence
if and only if $\pi[j]=s[i]$.

Geometrically, $\pi$ has an occurrence of $\sigma$ if there exists
a set of points in $\pi$ that is isomorphic to the set of points of $\sigma$.
In other words,
if there exists a set of points in $\pi$ with the same disposition as the set of points of $\sigma$,
regardless of the distance (see Figure \ref{example: pattern matching}).

@@

A $\textit{right-to-left maximum}$ (abbreviated to RLMax) of $\pi$ is an element that does not have any element above it and to its right (see Figure $\ref{fig:left to right maxima}$). Formally, $\pi[i]$ is a RLMax if and only if $\pi[i]$ is topmost element of $\pi[i:]$. Similarly $\pi[i]$ is a $\textit{right-to-left minimum}$ (abbreviate RLMin) if and only if $\pi[i]$ is the bottommost element of $\pi[i:]$.

@@

	\caption[Example pattern matching]{
		The element is a RLMax if and only if the hashed area is empty.}
	\label{fig:left to right maxima}
\end{figure}

@@

Geometrically, we represent underlined and overlined elements by \textit{forbidden areas}.
If two elements are overlined then we draw a horizontal forbidden area between those two points.
In an occurrence, we also draw the forbidden area between the matching of two overlined elements.
This area must be empty to have a valid occurrence of the bivincular pattern.
If the area is empty then there is no element between them when reading the permutation
from bottom to top,
in other words the matching elements are consecutive in value.
If two elements are underlined then we draw an vertical forbidden area.
(See Figure \ref{example:bivincular pattern matching}).

@@

\begin{corollary}
\label{corollary:minmaxelement}
$\pi$ is a wedge permutation of size $n$ if and only if for $1 \leq i \leq n$,
$\pi[i]$ is a RLMax or a RLMin.
\end{corollary}

\begin{proof} %[of Lemma~\ref{lemma:first element is 1 or n}]
Given a position $i$, $\pi[i:]$ is a wedge permutation,
so it either starts with the maximal or the minimal element.
\qed
\end{proof}

As a consequence, every element of a wedge permutation can be partitioned
into three sets: the set of RLMax elements, the set of RLMin elements and the
rightmost element which can be both a RLMax element and a RLMin element.

\begin{remark}
The set of RLMax elements is a decreasing subsequence,
indeed each element is left above the next RLMax element by definition
of a RLMax element. In the same fashion,
the set of RLMin elements is an increasing subsequence.
See Figure~\ref{fig:shape of the permutation plus factor}.
\end{remark}

\begin{remark}
Every element of the set of RLMax elements is above the elements
of the set of the RLMin elements.
In other words, a wedge permutations
can be cut in half horizontally on the rightmost element such that the upper half contains a decreasing
subsequence and the lower half contains an increasing subsequence.
This shape the permutation as a $>$,
hence the name of wedge permutations.
See Figure~\ref{fig:shape of the permutation plus factor}.
\end{remark}

\begin{remark}
The partition gives a bijection between the set of wedge permutations of size $n$ where the elements are $1,\ldots,n$
and the set of binary word of size $n-1$.
The word $w$ which corresponds to $\pi$ is the word where each letter at position $i$
represents whether $\pi[i]$ is a RLMax element or a RLMin element.
We call this bijection $\bijection$.
\end{remark}

To figure out whether or not an element is a RLMax element or a RLMin element,
one has to run the whole permutation from left to right starting
from the position of the element, we give a corollary of lemma~\ref{lemma:first element is 1 or n}, to guest it in constant time.

\begin{corollary}
\label{corollary:max is ascent}
Let $\pi$ be a wedge permutation and $1 \leq i < n$. Then,
\begin{enumerate}
\item $\pi[i]$ is a RLMin element if and only if $\pi[i]<\pi[i+1]$
\item $\pi[i]$ is a RLMax element if and only if $\pi[i]>\pi[i+1]$ is a RLMax
\end{enumerate}

@@

\begin{proof} %[of Lemma~\ref{lemma:MatchStripeToPermutation}]
The forward direction is obvious.
We prove the backward direction by induction on the size of
$\sigma$: if $\bijection(t)=\bijection(\sigma)$ then $t$ is an occurrence of $\sigma$.
The base case is a pattern of size $2$.
Suppose that $\sigma = 12$ and thus $\bijection(\sigma) = RLMin$.
Let $t = \pi_{i_1}\pi_{i_2}$, $i_1 < i_2$, be a subsequence of $\pi$
such that $\bijection(t) = RLMin$, this reduces to saying that
$\pi_{i_1} < \pi_{i_2}$,
and hence that $t$ is an occurrence of $\sigma = 12$ in $\pi$.
A similar argument shows that the lemma holds for $\sigma = 21$.
Now, assume that the lemma is true for all patterns up to size $k \geq 2$.
Let $\sigma$ be a wedge permutation of size $k+1$ and
let $t$
be a subsequence of $\pi$ of size $k+1$ such that
$\bijection(t) = \bijection(\sigma)$.
As $\bijection(t)[2:] = \bijection(\sigma)[2:]$
by the induction hypothesis, it follows that
$t[2:]$ is an occurrence of $\sigma[2:]$.
Moreover $\bijection(t)[1] = \bijection(\sigma)[1]$
thus $t[1]$ and $\sigma[1]$ are both either the bottommost or the topmost
element of their respective sequences.
Therefore, $t$ is an occurrence of $\sigma$ in $\pi$.
\qed
\end{proof}

@@

This section focuses on the permutation pattern matching problem
in case only the pattern $\sigma$ is a wedge permutation.
We need to consider a specific decomposition of $\sigma$ into \textit{factors}:
we split the permutation into maximal sequences of consecutive RLMin and RLMax elements, respectively called a RLMin factor and a RLMax factor.
This corresponds to splitting the permutation between every pair of RLMin-RLMax and RLMax-RLMin elements (see Figure \ref{fig:shape of the permutation plus factor}).
For the special case of a wedge permutation, this also corresponds to split the permutation into maximal sequences of elements consecutive in value.
We label the factors from right to left.
Note that the rightmost element can be both a RLMin or a RLMax,
we take the convention that it is as the same type as the element before it for the factorization,
such that the first factor (the rightmost factor) always contains at least two
elements.
For example,
$\sigma = 123984765$ is split as
$123-98-4-765$.
Hence
$\sigma =$ $\factor(4)$ $\factor(3)$ $\factor(2)$ $\factor(1)$ with
$\factor(4) =123$, $\factor(3) = 98$, $\factor(2) = 4$ and $\factor(1) = 765$.

@@

\begin{remark}
A rectangle contains
an occurrence of a RLMin (resp. RLMax) factor if and only if
the rectangle contains an increasing (resp. decreasing) subsequence
of same size or larger than the size of the factor.
\end{remark}

@@

Remark that by modifying the definition of $\PM^\pi_\sigma(i,i',((j,lb),(n,up)))$
we can reduce the computational time of the algorithm.
Instead of asking for any occurrence in $((j,lb),(n,up)))$,
we ask for the occurrences that start at $\pi[j]$.
As any element in the pattern is a RLMax or a RLMin,
the value of $\pi[j]$ can be used has a top or a bottom edge
of a rectangle.
So we can remove one argument from the problem,
reducing the maximal number of different problems that we have to compute.
%The new definition gives us:
%$\pi$ contains an occurrence of $\sigma$
%if and only if $\bigvee_{1 \leq j \leq n}  \PM^\pi_\sigma(\ell,1,((j,1),(n,n)))$
%is true
%where $\ell$ is the number of factors in $\sigma$.
\qed
\end{proof}

We extend our research to reduce the number of pairs of rectangles
that we have to test.
Informally,
given that the leftmost factor is a RLMin (resp. RLMax) factor and
given that the left top (resp. bottom left) corner of the second rectangle is fixed and
that the second rectangle contains the occurrence of the
rest of the permutation, it is in the interest of finding an occurrence in the whole rectangle
that the bottom (resp. top) edge of the second rectangle is the highest (resp. lowest) possible.
Indeed, the first rectangle will have
a highest top (resp. a lowest bottom) edge so it will contains more elements
which is better to find an occurrence.
The next lemma indicates which element is the topmost and the bottommost,
and the next one and its corollary formalise what said above.
But first we introduce another notation.

@@

\begin{lemma}
\label{corollary:whereIsMax}
Given a wedge permutation $\sigma$,
if $\factor(i)$ is a RLMin (resp. RLMax) factor
the topmost (resp. bottommost) element of
$\factor(i)\ldots \factor(1)$
is the leftmost element of $\factor(i-1)$.
\end{lemma}

\begin{proof} %[of Corollary~\ref{corollary:whereIsMax}]
This corollary states that, given a wedge permutation
if the permutation starts with a RLMin (resp. RLMax) element then the topmost (resp. bottommost) element of this permutation is the first RLMax (resp. RLMin) element (see Figure \ref{fig:where to find the max}). This is easy to see from the shape of a wedge permutation.
Formally, the RLMin elements are below the RLMax element, thus the topmost element must be
the first RLMax element, which is the leftmost element of $\factor(i-1)$.
\qed
\end{proof}

@@

\begin{lemma}
\label{lemma:ts}
Let $\ppattern$ be a wedge permutation and
$\factor(i)$ be an RLMin (resp. RLMax) factor $\sigma$.
Let $\pi$ be a permutation and $s$ a subsequence of $\pi$ such that
$s \in \SET{\ppattern}{\ptext}{i}{j}$ and
$s$ minimizes (resp. maximizes) the matching of the leftmost element of $\factor(i-1)$.
Let $s'$ be a subsequence of $\pi$ such that
$s \in \SET{\ppattern}{\ptext}{i}{j}$
and let $t=t's'$ be a subsequence of $\pi$ that extends $s'$ on the right.
Assume $t$ is an occurrence of $\factor(i+1)\ldots \factor(1)$ such that the leftmost element of $\factor(i)$ is matched to $\ptext[j]$. Then the subsequence $t's$ is also an occurrence of $\factor(i+1)\ldots \factor(1)$ such that the leftmost element of $\factor(i)$ is matched to $\ptext[j]$.
\end{lemma}

Informally, this lemma states that we can replace
the rectangle of an occurrence in $\SET{\ppattern}{\ptext}{i}{j}$
by another rectangle of an occurrence in $\SET{\ppattern}{\ptext}{i}{j}$
such that the second rectangle shares the same edges as the first one,
except for the top edge which is lower.
Especially the second rectangle can be chosen as the one with
the lowest top edge in all the rectangles sharing the right, bottom
and left edges.
More formally, given any occurrence of $\factor(i+1)$ $\factor(i)$ $\ldots$ $\factor(1)$,
where $\factor(i)$ is a RLMin (resp. RLMax) factor,
we can replace the part of the occurrence where $\factor(i)$ $\ldots$ $\factor(1)$ occurs, by any occurrence
that minimises (resp. maximises) the leftmost element of $\factor(i-1)$. Indeed the leftmost element of $\factor(i-1)$ is the topmost (resp. bottommost) element of  $\factor(i)$ $\ldots$ $\factor(1)$ (see Figure \ref{fig:where to find the max}).

\begin{proof} %[of Lemma~\ref{lemma:ts}]
Let us consider the case where $\factor(i)$ is a RLMin factor.
By definition $s$ is an occurrence of $\ppattern[\LMEi(\factor(i)):]$.
First of all, remark that $t'$ is an occurrence of $\factor(i+1)$.
So to prove that $t's$ is an occurrence of $\factor(i+1)\ldots \factor(1)$ we need to prove that the elements of t' are above the elements of $s$.
Since $t's'$ is an occurrence of $\factor(i+1)\ldots \factor(1)$ it follows that the elements of $t'$ are above the elements of $s'$. Moreover the topmost element of $s$ is below (or equal to) the topmost element of $s'$ thus the elements of $s$ are below the elements of $t'$. We use a symmetric argument if $\factor(i)$ is a RLMax factor.
\qed
\end{proof}

\begin{corollary}
\label{corollary:we can chose a matching}
Let $\ppattern$ be a permutation,
$\factor(i)$ be a RLMin (resp. RLMax) factor
and
$s$ be a subsequence of $\pi$ such that $s \in \SET{\ppattern}{\ptext}{i}{j}$ and
that minimizes (resp. maximizes) the matching of the leftmost element of $\factor(i-1)$.
The following statements are equivalent:
\begin{itemize}
	\item There exists an occurrence
	 of $\ppattern$ in $\ptext$ where the leftmost element of $\factor(i)$ is matched to $\ptext[j]$.
	\item There exists an occurrence $t$ of $\factor(\ell)\ldots \factor(i+1)$ in $\ptext[:j-1]$  such that $ts$ is an occurrence of $\ppattern$ in $\ptext$ with the leftmost element of $\factor(i)$ is matched to $\ptext[j]$.
\end{itemize}
\end{corollary}

@@

\begin{proof}
The algorithm follows the previous one,
but instead of return true or false,
assuming that $\factor(i)$ is a RLMin (resp. RLMax) factor
the algorithm return the value of the
top (resp. bottom) edge.
So that once we compute the problem for $R_2$
we can use this value has the top (resp. bottom) edge
for the rectangle $R_1$.

@@

Given a RLMin (resp. RLMax) factor $\factor(i)$  of $\sigma$ and a position $j$ in $\pi$,
we want the value of the top (resp. bottom) edge
of any rectangle with left bottom (resp. with left top) corner   $(j,\pi[j])$  containing an occurrence of  $\factor(i) \ldots \factor(1)$ starting at $\pi[j]$
and which minimizes (resp. maximizes) its top (resp. bottom) edge
or a value which indicate that no occurrence exists in this rectangle. More formally:
\begin{itemize}
\item If $\factor(i)$ is RLMin factor.

@@

\item If $\factor(i)$ is RLMax factor.
$$
\PM^\pi_\sigma(i,j) =
$$
$$
\begin{cases}

@@

We show how to compute recursively those values.
\begin{itemize}
\item If the permutation is reduced to a RLMin (resp. RLMax) factor
then one has to decide the top (resp. lower ) edge of
a rectangle that
contains an increasing or a decreasing factor
with the same size of larger than the
size of the unique factor starting at $\pi[j]$.
This case happens when $i=1$:

\begin{itemize}
\item If $\factor(i)$ is RLMin factor.
$$
\PM^\pi_\sigma(1,j) = \min \text{\{$\pi[j']$ $ |$ if $ |factor(1)| \leq LIS_{\ptext}(j,j',\pi[j'])$\}}_{j'\geq j} \cup \{\infty\}
$$

\item If $\factor(i)$ is RLMax factor.
$$
\PM^\pi_\sigma(1,j) = \min \text{\{$\pi[j']$ $ |$ if $ |factor(1)| \leq LDS_{\ptext}(j,j',\pi[j'])$\}}_{j'\geq j} \cup \{0\}
$$
\end{itemize}

\item Otherwise, if $\factor(i)$ is a RLMin (resp. RLMax) factor,
 we split the rectangle into two rectangles $R_1$ and
$R_2$ such that $R_1$ is left below $R_2$,
$R_1$ contains an occurrence of $\factor(i)$ starting at
$j$
and $R_2$ contains an occurrence of $\factor(i-1) \ldots factor(1)$
starting at $j'>j$.
As said before, we only need to test the pair of rectangles
where the rectangle $R_2$ has for left edge $j'$
and has the highest bottom (resp. lowest top) edge.
So for $R_2$ we want
a rectangle starting at $j'$
which contains an occurrence of $\factor(i-1) \ldots factor(i')$
starting at $j'$
with maximize the bottom (resp. minimize the top) edge.
Moreover we also want to know the value of the bottom (resp. top) edge to use it as a bound to $R_1$ so that
that $R_1$ is below $R_2$,
in other words we want $\PM^\pi_\sigma(i-1,j')$.
As before $R_1$ has to be biggest possible and
to ensure that $R_1$ is left below (resp. above) $R_2$,
$R_1$ must have for top right corner $(j'-1, \PM^\pi_\sigma(i-1,j')-1 )$
(resp. bottom right corner $(j'-1, \PM^\pi_\sigma(i-1,j')+1 )$).
Finally in all the "correct" pairs of rectangles,
we need the value of the top (resp. bottom) edge of a rectangle minimizing the top edge (resp. maximizing the bottom edge)
containing the pair of rectangles. Note that the top (resp. bottom) edge has for value $\pi[j']$
(the matching of  $\LMEi(\factor(i-1))$). Formally:

\item If $\factor(i)$ is RLMin factor:\\
$$
\PM^\pi_\sigma(i,j) =
$$
$
\min \{\infty\} \cup \{ \pi[j'] \mid b=\PM^\pi_\sigma(i-1,j') \text{ is not $\infty$ and }
 |factor(i)| \leq LIS_{\ptext}(j,j'-1,b-1) \}_{j' < j}
$

\item If $\factor(i)$ is RLMax factor:\\
$$
\PM^\pi_\sigma(i,j) =
$$
$
\min \{\infty\} \cup \{ \pi[j'] \mid b=\PM^\pi_\sigma(i-1,j') \text{ is not $\infty$ and }
|factor(i)| \leq LDS_{\ptext}(j,j'-1,b+1) \}_{j' < j}
$

@@

\begin{lemma}
\label{lemma:after}
Let $\sigma$ be a wedge permutation.
\begin{itemize}
\item If $\sigma[i]$ is a RLMin element and is not the rightmost element, then $\sigma[i]+1$ is to the right of $\sigma[i]$.
\item If $\sigma[i]$ is a RLMax element and is not the rightmost element, then $\sigma[i]-1$ is to the right of $\sigma[i]$.
\end{itemize}
\end{lemma}

@@

\begin{lemma}
\label{lemma:ascentDescentAscent}
Let $\sigma$ be a wedge permutation.
\begin{itemize}
\item If $m$ and $m+1$ are both RLMin elements
then any element between $m$ and $m+1$ (if any)
is a RLMax element.
\item If $m$ and $m-1$ are both RLmax elements
then any element between $m$ and $m-1$ (if any)
is a RLMin element.
\end{itemize}

>>>>

\newcommand{\RLMin}{\text{RLMin}}
\newcommand{\RLMax}{\text{RLMax}}

@@

For an occurrence $s$ of $\sigma$ in $\pi$,
we say that $\pi[j]$ correspond $\sigma[i]$ in the occurrence or
that $\sigma[i]$ is matched to $\pi[j]$  in the occurrence
if and only if $\pi[j]=s[i]$.
Geometrically, $\pi$ has an occurrence of $\sigma$ if there exists
a set of points in $\pi$ whose relative positions are the same as those of $\sigma$
(see Figure \ref{example: pattern matching}).
%% In other words,
%% if there exists a set of points in $\pi$ with the same disposition as the set of points of $\sigma$,
%% regardless of the distance (see Figure \ref{example: pattern matching}).

@@

A $\textit{right-to-left maximum}$ (\RLMax, for short) of $\pi$ is an element that does not have any element above it and to its right (see Figure $\ref{fig:left to right maxima}$). Formally, $\pi[i]$ is a \RLMax if and only if $\pi[i]$ is topmost element of $\pi[i:]$. Similarly $\pi[i]$ is a $\textit{right-to-left minimum}$ (\RLMin, for short) if and only if $\pi[i]$ is the bottommost element of $\pi[i:]$.

@@

	\caption[Example pattern matching]{
		The element is a \RLMax if and only if the hashed area is empty.}
	\label{fig:left to right maxima}
\end{figure}

@@

Geometrically, we represent underlined and overlined elements by \textit{forbidden areas}.
If two elements are overlined then we draw a horizontal forbidden area between those two points.
In an occurrence, we also draw the forbidden area between the matching of two overlined elements.
This area must be empty to have a valid occurrence of the bivincular pattern.
If the area is empty then there is no element between them when reading the permutation
from bottom to top,
in other words the matching elements are consecutive in value.
If two elements are underlined then we draw a vertical forbidden area.
(See Figure \ref{example:bivincular pattern matching}).

@@

\begin{corollary}
\label{corollary:minmaxelement}
$\pi$ is a wedge permutation of size $n$ if and only if for $1 \leq i \leq n$,
$\pi[i]$ is a \RLMax or a \RLMin.
\end{corollary}

%% \begin{proof} %[of Lemma~\ref{lemma:first element is 1 or n}]
%% Given a position $i$, $\pi[i:]$ is a wedge permutation,
%% so it either starts with the maximal or the minimal element.
%% \qed
%% \end{proof}

As a consequence, every element of a wedge permutation can be partitioned
into three sets: the set of \RLMax elements, the set of \RLMin elements and the
rightmost element which can be both a \RLMax element and a \RLMin element.

\begin{remark}
The set of \RLMax elements is a decreasing subsequence,
indeed each element is left above the next \RLMax element by definition
of a \RLMax element. In the same fashion,
the set of \RLMin elements is an increasing subsequence.
See Figure~\ref{fig:shape of the permutation plus factor}.
\end{remark}

\begin{remark}
Every element of the set of \RLMax elements is above the elements
of the set of the \RLMin elements.
In other words, a wedge permutations
can be cut in half horizontally on the rightmost element such that the upper half contains a decreasing
subsequence and the lower half contains an increasing subsequence.
This shape the permutation as a $>$,
hence the name of wedge permutations.
See Figure~\ref{fig:shape of the permutation plus factor}.
\end{remark}

\begin{remark}
The partition gives a bijection between the set of wedge permutations of size $n$ where the elements are $1,\ldots,n$
and the set of binary word of size $n-1$.
The word $w$ which corresponds to $\pi$ is the word where each letter at position $i$
represents whether $\pi[i]$ is a \RLMax element or a \RLMin element.
We call this bijection $\bijection$.
\end{remark}

To figure out whether or not an element is a \RLMax element or a \RLMin element,
one has to run the whole permutation from left to right starting
from the position of the element, we give a corollary of lemma~\ref{lemma:first element is 1 or n}, to guest it in constant time.

\begin{corollary}
\label{corollary:max is ascent}
Let $\pi$ be a wedge permutation and $1 \leq i < n$. Then,
\begin{enumerate}
\item $\pi[i]$ is a \RLMin element if and only if $\pi[i]<\pi[i+1]$
\item $\pi[i]$ is a \RLMax element if and only if $\pi[i]>\pi[i+1]$ is a \RLMax
\end{enumerate}

@@

\begin{proof} %[of Lemma~\ref{lemma:MatchStripeToPermutation}]
The forward direction is obvious.
We prove the backward direction by induction on the size of
$\sigma$: if $\bijection(t)=\bijection(\sigma)$ then $t$ is an occurrence of $\sigma$.
The base case is a pattern of size $2$.
Suppose that $\sigma = 12$ and thus $\bijection(\sigma) = \RLMin$.
Let $t = \pi_{i_1}\pi_{i_2}$, $i_1 < i_2$, be a subsequence of $\pi$
such that $\bijection(t) = \RLMin$, this reduces to saying that
$\pi_{i_1} < \pi_{i_2}$,
and hence that $t$ is an occurrence of $\sigma = 12$ in $\pi$.
A similar argument shows that the lemma holds for $\sigma = 21$.
Now, assume that the lemma is true for all patterns up to size $k \geq 2$.
Let $\sigma$ be a wedge permutation of size $k+1$ and
let $t$
be a subsequence of $\pi$ of size $k+1$ such that
$\bijection(t) = \bijection(\sigma)$.
As $\bijection(t)[2:] = \bijection(\sigma)[2:]$
by the induction hypothesis, it follows that
$t[2:]$ is an occurrence of $\sigma[2:]$.
Moreover $\bijection(t)[1] = \bijection(\sigma)[1]$
thus $t[1]$ and $\sigma[1]$ are both either the bottommost or the topmost
element of their respective sequences.
Therefore, $t$ is an occurrence of $\sigma$ in $\pi$.
\qed
\end{proof}

@@

This section focuses on the permutation pattern matching problem
in case only the pattern $\sigma$ is a wedge permutation.
We need to consider a specific decomposition of $\sigma$ into \textit{factors}:
we split the permutation into maximal sequences of consecutive \RLMin and \RLMax elements, respectively called a \RLMin factor and a \RLMax factor.
This corresponds to splitting the permutation between every pair of \RLMin-\RLMax and \RLMax-\RLMin elements (see Figure \ref{fig:shape of the permutation plus factor}).
For the special case of a wedge permutation, this also corresponds to split the permutation into maximal sequences of elements consecutive in value.
We label the factors from right to left.
Note that the rightmost element can be both a \RLMin or a \RLMax,
we take the convention that it is as the same type as the element before it for the factorization,
such that the first factor (the rightmost factor) always contains at least two
elements.
For example,
$\sigma = 123984765$ is split as
$123-98-4-765$.
Hence
$\sigma =$ $\factor(4)$ $\factor(3)$ $\factor(2)$ $\factor(1)$ with
$\factor(4) =123$, $\factor(3) = 98$, $\factor(2) = 4$ and $\factor(1) = 765$.

@@

\begin{remark}
A rectangle contains
an occurrence of a \RLMin (resp. \RLMax) factor if and only if
the rectangle contains an increasing (resp. decreasing) subsequence
of same size or larger than the size of the factor.
\end{remark}

@@

Remark that by modifying the definition of $\PM^\pi_\sigma(i,i',((j,lb),(n,up)))$
we can reduce the computational time of the algorithm.
Instead of asking for any occurrence in $((j,lb),(n,up)))$,
we ask for the occurrences that start at $\pi[j]$.
As any element in the pattern is a \RLMax or a \RLMin,
the value of $\pi[j]$ can be used has a top or a bottom edge
of a rectangle.
So we can remove one argument from the problem,
reducing the maximal number of different problems that we have to compute.
%The new definition gives us:
%$\pi$ contains an occurrence of $\sigma$
%if and only if $\bigvee_{1 \leq j \leq n}  \PM^\pi_\sigma(\ell,1,((j,1),(n,n)))$
%is true
%where $\ell$ is the number of factors in $\sigma$.
\qed
\end{proof}

We extend our research to reduce the number of pairs of rectangles
that we have to test.
Informally,
given that the leftmost factor is a \RLMin (resp. \RLMax) factor and
given that the left top (resp. bottom left) corner of the second rectangle is fixed and
that the second rectangle contains the occurrence of the
rest of the permutation, it is in the interest of finding an occurrence in the whole rectangle
that the bottom (resp. top) edge of the second rectangle is the highest (resp. lowest) possible.
Indeed, the first rectangle will have
a highest top (resp. a lowest bottom) edge so it will contains more elements
which is better to find an occurrence.
The next lemma indicates which element is the topmost and the bottommost,
and the next one and its corollary formalise what said above.
But first we introduce another notation.

@@

\begin{lemma}
\label{corollary:whereIsMax}
Given a wedge permutation $\sigma$,
if $\factor(i)$ is a \RLMin (resp. \RLMax) factor
the topmost (resp. bottommost) element of
$\factor(i)\ldots \factor(1)$
is the leftmost element of $\factor(i-1)$.
\end{lemma}

\begin{proof} %[of Corollary~\ref{corollary:whereIsMax}]
This corollary states that, given a wedge permutation
if the permutation starts with a \RLMin (resp. \RLMax) element then the topmost (resp. bottommost) element of this permutation is the first \RLMax (resp. \RLMin) element (see Figure \ref{fig:where to find the max}). This is easy to see from the shape of a wedge permutation.
Formally, the \RLMin elements are below the \RLMax element, thus the topmost element must be
the first \RLMax element, which is the leftmost element of $\factor(i-1)$.
\qed
\end{proof}

@@

\begin{lemma}
\label{lemma:ts}
Let $\ppattern$ be a wedge permutation and
$\factor(i)$ be an \RLMin (resp. \RLMax) factor $\sigma$.
Let $\pi$ be a permutation and $s$ a subsequence of $\pi$ such that
$s \in \SET{\ppattern}{\ptext}{i}{j}$ and
$s$ minimizes (resp. maximizes) the matching of the leftmost element of $\factor(i-1)$.
Let $s'$ be a subsequence of $\pi$ such that
$s \in \SET{\ppattern}{\ptext}{i}{j}$
and let $t=t's'$ be a subsequence of $\pi$ that extends $s'$ on the right.
Assume $t$ is an occurrence of $\factor(i+1)\ldots \factor(1)$ such that the leftmost element of $\factor(i)$ is matched to $\ptext[j]$. Then the subsequence $t's$ is also an occurrence of $\factor(i+1)\ldots \factor(1)$ such that the leftmost element of $\factor(i)$ is matched to $\ptext[j]$.
\end{lemma}

Informally, this lemma states that we can replace
the rectangle of an occurrence in $\SET{\ppattern}{\ptext}{i}{j}$
by another rectangle of an occurrence in $\SET{\ppattern}{\ptext}{i}{j}$
such that the second rectangle shares the same edges as the first one,
except for the top edge which is lower.
Especially the second rectangle can be chosen as the one with
the lowest top edge in all the rectangles sharing the right, bottom
and left edges.
More formally, given any occurrence of $\factor(i+1)$ $\factor(i)$ $\ldots$ $\factor(1)$,
where $\factor(i)$ is a \RLMin (resp. \RLMax) factor,
we can replace the part of the occurrence where $\factor(i)$ $\ldots$ $\factor(1)$ occurs, by any occurrence
that minimises (resp. maximises) the leftmost element of $\factor(i-1)$. Indeed the leftmost element of $\factor(i-1)$ is the topmost (resp. bottommost) element of  $\factor(i)$ $\ldots$ $\factor(1)$ (see Figure \ref{fig:where to find the max}).

\begin{proof} %[of Lemma~\ref{lemma:ts}]
Let us consider the case where $\factor(i)$ is a \RLMin factor.
By definition $s$ is an occurrence of $\ppattern[\LMEi(\factor(i)):]$.
First of all, remark that $t'$ is an occurrence of $\factor(i+1)$.
So to prove that $t's$ is an occurrence of $\factor(i+1)\ldots \factor(1)$ we need to prove that the elements of t' are above the elements of $s$.
Since $t's'$ is an occurrence of $\factor(i+1)\ldots \factor(1)$ it follows that the elements of $t'$ are above the elements of $s'$. Moreover the topmost element of $s$ is below (or equal to) the topmost element of $s'$ thus the elements of $s$ are below the elements of $t'$. We use a symmetric argument if $\factor(i)$ is a \RLMax factor.
\qed
\end{proof}

\begin{corollary}
\label{corollary:we can chose a matching}
Let $\ppattern$ be a permutation,
$\factor(i)$ be a \RLMin (resp. \RLMax) factor
and
$s$ be a subsequence of $\pi$ such that $s \in \SET{\ppattern}{\ptext}{i}{j}$ and
that minimizes (resp. maximizes) the matching of the leftmost element of $\factor(i-1)$.
The following statements are equivalent:
\begin{itemize}
	\item There exists an occurrence
	 of $\ppattern$ in $\ptext$ where the leftmost element of $\factor(i)$ is matched to $\ptext[j]$.
	\item There exists an occurrence $t$ of $\factor(\ell)\ldots \factor(i+1)$ in $\ptext[:j-1]$  such that $ts$ is an occurrence of $\ppattern$ in $\ptext$ with the leftmost element of $\factor(i)$ is matched to $\ptext[j]$.
\end{itemize}
\end{corollary}

@@

\begin{proof}
The algorithm follows the previous one,
but instead of return true or false,
assuming that $\factor(i)$ is a \RLMin (resp. \RLMax) factor
the algorithm return the value of the
top (resp. bottom) edge.
So that once we compute the problem for $R_2$
we can use this value has the top (resp. bottom) edge
for the rectangle $R_1$.

@@

Given a \RLMin (resp. \RLMax) factor $\factor(i)$  of $\sigma$ and a position $j$ in $\pi$,
we want the value of the top (resp. bottom) edge
of any rectangle with left bottom (resp. with left top) corner   $(j,\pi[j])$  containing an occurrence of  $\factor(i) \ldots \factor(1)$ starting at $\pi[j]$
and which minimizes (resp. maximizes) its top (resp. bottom) edge
or a value which indicate that no occurrence exists in this rectangle. More formally:
\begin{itemize}
\item If $\factor(i)$ is \RLMin factor.

@@

\item If $\factor(i)$ is \RLMax factor.
$$
\PM^\pi_\sigma(i,j) =
$$
$$
\begin{cases}

@@

We show how to compute recursively those values.
\begin{itemize}
\item If the permutation is reduced to a \RLMin (resp. \RLMax) factor
then one has to decide the top (resp. lower ) edge of
a rectangle that
contains an increasing or a decreasing factor
with the same size of larger than the
size of the unique factor starting at $\pi[j]$.
This case happens when $i=1$:

\begin{itemize}
\item If $\factor(i)$ is \RLMin factor.
$$
\PM^\pi_\sigma(1,j) = \min \text{\{$\pi[j']$ $ |$ if $ |factor(1)| \leq LIS_{\ptext}(j,j',\pi[j'])$\}}_{j'\geq j} \cup \{\infty\}
$$

\item If $\factor(i)$ is \RLMax factor.
$$
\PM^\pi_\sigma(1,j) = \min \text{\{$\pi[j']$ $ |$ if $ |factor(1)| \leq LDS_{\ptext}(j,j',\pi[j'])$\}}_{j'\geq j} \cup \{0\}
$$
\end{itemize}

\item Otherwise, if $\factor(i)$ is a \RLMin (resp. \RLMax) factor,
 we split the rectangle into two rectangles $R_1$ and
$R_2$ such that $R_1$ is left below $R_2$,
$R_1$ contains an occurrence of $\factor(i)$ starting at
$j$
and $R_2$ contains an occurrence of $\factor(i-1) \ldots factor(1)$
starting at $j'>j$.
As said before, we only need to test the pair of rectangles
where the rectangle $R_2$ has for left edge $j'$
and has the highest bottom (resp. lowest top) edge.
So for $R_2$ we want
a rectangle starting at $j'$
which contains an occurrence of $\factor(i-1) \ldots factor(i')$
starting at $j'$
with maximize the bottom (resp. minimize the top) edge.
Moreover we also want to know the value of the bottom (resp. top) edge to use it as a bound to $R_1$ so that
that $R_1$ is below $R_2$,
in other words we want $\PM^\pi_\sigma(i-1,j')$.
As before $R_1$ has to be biggest possible and
to ensure that $R_1$ is left below (resp. above) $R_2$,
$R_1$ must have for top right corner $(j'-1, \PM^\pi_\sigma(i-1,j')-1 )$
(resp. bottom right corner $(j'-1, \PM^\pi_\sigma(i-1,j')+1 )$).
Finally in all the "correct" pairs of rectangles,
we need the value of the top (resp. bottom) edge of a rectangle minimizing the top edge (resp. maximizing the bottom edge)
containing the pair of rectangles. Note that the top (resp. bottom) edge has for value $\pi[j']$
(the matching of  $\LMEi(\factor(i-1))$). Formally:

\item If $\factor(i)$ is \RLMin factor:\\
$$
\PM^\pi_\sigma(i,j) =
$$
$
\min \{\infty\} \cup \{ \pi[j'] \mid b=\PM^\pi_\sigma(i-1,j') \text{ is not $\infty$ and }
 |factor(i)| \leq LIS_{\ptext}(j,j'-1,b-1) \}_{j' < j}
$

\item If $\factor(i)$ is \RLMax factor:\\
$$
\PM^\pi_\sigma(i,j) =
$$
$
\min \{\infty\} \cup \{ \pi[j'] \mid b=\PM^\pi_\sigma(i-1,j') \text{ is not $\infty$ and }
|factor(i)| \leq LDS_{\ptext}(j,j'-1,b+1) \}_{j' < j}
$

@@

\begin{lemma}
\label{lemma:after}
Let $\sigma$ be a wedge permutation.
\begin{itemize}
\item If $\sigma[i]$ is a \RLMin element and is not the rightmost element, then $\sigma[i]+1$ is to the right of $\sigma[i]$.
\item If $\sigma[i]$ is a \RLMax element and is not the rightmost element, then $\sigma[i]-1$ is to the right of $\sigma[i]$.
\end{itemize}
\end{lemma}

@@

\begin{lemma}
\label{lemma:ascentDescentAscent}
Let $\sigma$ be a wedge permutation.
\begin{itemize}
\item If $m$ and $m+1$ are both \RLMin elements
then any element between $m$ and $m+1$ (if any)
is a \RLMax element.
\item If $m$ and $m-1$ are both RLmax elements
then any element between $m$ and $m-1$ (if any)
is a \RLMin element.
\end{itemize}
